How to Not write Virtual Machines for Dynamic Languages

### 2007-05-26 talk/dyla2007/dyla.tex
fix some of the references

\item
this meta-programming component of PyPy is called the \emph{translation
framework,} as it translates RPython source code (i.e. the full Python
interpreter) into lower-level code.  Its purpose is to add aspects to
and specialize the interpreter to fit a selectable virtual or hardware
runtime environment.  This either turns the interpreter into a
standalone virtual machine, or integrates it into an existing OO VM.
The necessary support code - e.g. the garbage collector when
targeting C - is itself written in RPython in much the same spirit
that the Jikes RVM's GCs are written in Java [JIKES]; as needed, it is
translated together with the interpreter to form the final custom VM.
\end{itemize}

A detailed description of this translation process is beyond the scope of the
present paper; it can be found in \cite{pypyvmconstruction}.  The actual Python
interpreter of PyPy and the results we achieved by translating it to C, LLVM
\ref{LLVM:CGO04} and .NET are described in [XXX].  These results show that the
approach is practical and gives results whose performance is within the same
order of magnitude (within a factor of 2 and improving) of the hand-written,
well-tuned CPython, the C reference implementation.

@@

A first point is that it makes interpreters easy to write, update and
generally experiment with.  More expressiveness helps at all levels: our
Python interpreter is written in RPython as a relatively simple
interpreter, in some respects easier to understand than CPython.  We are
using its high level and flexibility to quickly experiment with features
or implementation techniques in ways that would, in a traditional
approach, require pervasive changes to the source code.  For example,
PyPy's Python interpreter can optionally provide lazily computed objects
- a 150-lines extension in PyPy that would require global changes in
CPython.  Further examples can be found in our technical reports; we
should notably mention an extension adding a state-of-the-art security
model for Python based on data flow tacking [XXX], and general
performance improvements found by extensive experimentation [XXX], some
of which were back-ported to CPython.

@@

In fact, there is a well-known example of a CPython fork: Stackless
Python [XXX], which adds support for coroutines.  This is very hard to
do in CPython because the interpreter is written in a highly recursive
style.  Stackless Python required large-scale changes; it is not merged
back into CPython due to the pervasive increase in complexity that it
requires.  In PyPy, though, an optional "stackless" transformation is
able to turn the Python interpreter - also written in a simple highly
recursive style - into an efficient variant of continuation-passing
style (CPS), enabling the usage of coroutines in the translated
interpreter.  For more details and other examples of translation-level
transformations, see [XXX].

>>>>

\item
this meta-programming component of PyPy is called the \emph{translation
framework}, as it translates RPython source code (i.e. the full Python
interpreter) into lower-level code.  Its purpose is to add aspects to
and specialize the interpreter to fit a selectable virtual or hardware
runtime environment.  This either turns the interpreter into a
standalone virtual machine, or integrates it into an existing OO VM.
The necessary support code - e.g. the garbage collector when
targeting C - is itself written in RPython in much the same spirit
that the Jikes RVM's GCs are written in Java \cite{JikesGC}; as needed, it is
translated together with the interpreter to form the final custom VM.
\end{itemize}

A detailed description of this translation process is beyond the scope of the
present paper; it can be found in \cite{pypyvmconstruction}.  The actual Python
interpreter of PyPy and the results we achieved by translating it to C, LLVM
\cite{LLVM} and .NET are described in [XXX].  These results show that the
approach is practical and gives results whose performance is within the same
order of magnitude (within a factor of 2 and improving) of the hand-written,
well-tuned CPython, the C reference implementation.

@@

A first point is that it makes interpreters easy to write, update and
generally experiment with.  More expressiveness helps at all levels: our
Python interpreter is written in RPython as a relatively simple
interpreter, in some respects easier to understand than CPython.  We are
using its high level and flexibility to quickly experiment with features
or implementation techniques in ways that would, in a traditional
approach, require pervasive changes to the source code.  For example,
PyPy's Python interpreter can optionally provide lazily computed objects
- a 150-lines extension in PyPy that would require global changes in
CPython.  Further examples can be found in our technical reports; we
should notably mention an extension adding a state-of-the-art security
model for Python based on data flow tacking \cite{D12.1}, and general
performance improvements found by extensive experimentation \cite{D06.1}, some
of which were back-ported to CPython.

@@

In fact, there is a well-known example of a CPython fork: Stackless
Python \cite{stackless}, which adds support for coroutines.  This is very hard to
do in CPython because the interpreter is written in a highly recursive
style.  Stackless Python required large-scale changes; it is not merged
back into CPython due to the pervasive increase in complexity that it
requires.  In PyPy, though, an optional "stackless" transformation is
able to turn the Python interpreter - also written in a simple highly
recursive style - into an efficient variant of continuation-passing
style (CPS), enabling the usage of coroutines in the translated
interpreter.  For more details and other examples of translation-level
transformations, see \cite{D07.1}.
